How is experience transformed into memory?

### 2018-09-02 paper/boneyard.tex
backing up progress (results)

>>>>

\subsection*{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters present, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text). As depicted in Fig.~\ref{fig:model}A, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}C).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly discriminable. We found that the ``narrative details'' (a few sentences summarizing the dialogue and specific happenings of the scene) of the video were the most important feature driving this model structure (see Fig.~\featureimportance A).

After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}B. Notably, participant \#13's recall model appears visually similar to the video model. Similar to the video model analysis (Fig.~\featureimportance A), we found that the ``narrative details'' feature was most important for driving the video/recall relationship (Fig.~\featureimportance B).  Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a sentences (68 to 294) by sentences (68 to 294) correlation matrix (Fig.~\ref{fig:model}D). Like the video model, each participant's recall correlation matrix exhibited a strong block-diagonal structure (Fig.~\corrmats). Notably, this suggests that participants recounted the video in discriminable segments, likely related to the recall of specific events from the video.

\subsection*{Segmenting the video and recall models into ``events''}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}C,D).  We hypothesized that this structure might arise from transient stability in the "theme" of both the video and of participants' memory for the video. We hypothesized that specific events described by a participant could be ``matched'' to specific video events by computing the similarity (i.e., correlation) between their topic vectors. In other words, the video event that is most thematically similar to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e., states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the video model and a representative participant's recall model are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}C,D).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}E).  We performed the same procedure for the recall matrices (Fig.~\ref{fig:model}F). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}G for example, Fig.~\matchmats for all participants). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we found the index of the video event with the highest correlation to the recall event (i.e., the argmax).  This is depicted in Fig.~\ref{fig:model}G as the cells highlighted with a yellow border. Notably, our algorithm suggests that the example participant recalled most of the video events and did so in order.

Then, we computed a group-averaged recall event model and video-recall event correlation matrix. For each participant (and each recall event), we sorted the recall event vectors (across all participants) according to the video event with the highest correlation. We then averaged the recall event vectors within each group. This yielded an average recall event vector for all but one (of 34) video events, since no participant remembered one of the events according to our model. Lastly, we computed an average recall event (34) by video event (34) correlation matrix, and highlighted the cell with the highest correlation value with a yellow border (Fig.~\ref{fig:model}H). Notably, this matrix displayed high correlation values along the diagonal and low correlations in the off-diagonal cells. This suggests that on average, participants were able to recapitulate the events in the episode in a specific, sequential and highly descriminable way.

\subsection*{The trajectory of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled \citep{Murd62a}. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization \citep{Kaha96, PolyEtal09}. However, these approaches cannot (and were not designed to) capture the rich temporal dynamics inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

%%%%

### 2018-09-02 paper/main.tex
backing up progress (results)

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (e.g.\ every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  For example, some participants' recall topic proportions segmented into just a few events (e.g.\ Participants 1, 4, and 15), while others' recalls segmented into many shorter duration events (e.g.\ Participants 12, 13, and 17).  This suggests that different participants may be recalling the video with different levels of detail-- e.g. some might touch on just the major plot points, whereas others might attempt to recall every minor scene.  The second clear pattern present in every individual participant's recall correlation matrix is that, unlike in the video correlation matrix, there are substantial off-diagonal correlations in participant's recalls.  Whereas each event in the original video (was largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory participants appear to be integrating \textit{across} different events, blending elements of previously recalled and not-yet-recalled events into each newly recalled event (Figs.~\ref{fig:model}D, \corrmats).

@@

% characterize gist of movie and recalls
% x1. movie: fit topic models (add note on dynamic topic models).  fig 1: what do the topics reflect (top words from each topic), example with a few scenes.
% x2. notice block diagonal structure, rather than smooth transitions; motivates HMM approach.  Note about selecting K
% x3. question: do recalls exhibit same structure?  apply topic model to recalls.  note that by projecting everything into a common topic space, the thematic content of each moment of movie/recall can be compared, even though the mix of words used to describe any given event or moment may not be consistent across people (or across annotations/recall)
% x4. recalls exhibit blocky structure, but at different scales.  motivates fitting K separately by person
% 5. match up events (correlations)
% 6. now we can use classic memory analyses to examine structure of recall.  motivate and reference PFR, lag-CRP, and SPC.  but this does not tell us about structure in the *content* of what people remembered, or whether people's recalls are capturing the hierarchical temporal structure of the original movie.
% 7. topic trajectories as a new way of characterizing recall.  assess agreement using rayleigh test.  also examine shapes of each subject's trajectory to show match.  key insight: everyone captures the general shape, but at different "sampling resolutions"
% 8. we can also project back onto the original (word) features to understand more about what any particular part of the trajectory reflects (wordle fig)
% 9. the above analyses tell us about how the trajectory of thematic content in the episode is both preserved and transformed in participants' memories.  we next carried out analyses to understand the neural bases of these phenomena.  analysis 1: which structures are sensitive (during viewing) to the original episode's topic trajectory?  answer: default mode.  anlaysis 2: which structure are sensitive (during viewing) to how the participants will later recall the episode?  answer: vmPFC, default mode.  (potential role of vmPFC in reflecting participant-specific memory transformations and processing)

\subsection*{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters present, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text). As depicted in Fig.~\ref{fig:model}A, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}C).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly discriminable. We found that the ``narrative details'' (a few sentences summarizing the dialogue and specific happenings of the scene) of the video were the most important feature driving this model structure (see Fig.~\featureimportance A).

After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}B. Notably, participant \#13's recall model appears visually similar to the video model. Similar to the video model analysis (Fig.~\featureimportance A), we found that the ``narrative details'' feature was most important for driving the video/recall relationship (Fig.~\featureimportance B).  Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a sentences (68 to 294) by sentences (68 to 294) correlation matrix (Fig.~\ref{fig:model}D). Like the video model, each participant's recall correlation matrix exhibited a strong block-diagonal structure (Fig.~\corrmats). Notably, this suggests that participants recounted the video in discriminable segments, likely related to the recall of specific events from the video.

\subsection*{Segmenting the video and recall models into ``events''}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}C,D).  We hypothesized that this structure might arise from transient stability in the "theme" of both the video and of participants' memory for the video. We hypothesized that specific events described by a participant could be ``matched'' to specific video events by computing the similarity (i.e., correlation) between their topic vectors. In other words, the video event that is most thematically similar to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e., states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the video model and a representative participant's recall model are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}C,D).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}E).  We performed the same procedure for the recall matrices (Fig.~\ref{fig:model}F). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}G for example, Fig.~\matchmats for all participants). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we found the index of the video event with the highest correlation to the recall event (i.e., the argmax).  This is depicted in Fig.~\ref{fig:model}G as the cells highlighted with a yellow border. Notably, our algorithm suggests that the example participant recalled most of the video events and did so in order.

Then, we computed a group-averaged recall event model and video-recall event correlation matrix. For each participant (and each recall event), we sorted the recall event vectors (across all participants) according to the video event with the highest correlation. We then averaged the recall event vectors within each group. This yielded an average recall event vector for all but one (of 34) video events, since no participant remembered one of the events according to our model. Lastly, we computed an average recall event (34) by video event (34) correlation matrix, and highlighted the cell with the highest correlation value with a yellow border (Fig.~\ref{fig:model}H). Notably, this matrix displayed high correlation values along the diagonal and low correlations in the off-diagonal cells. This suggests that on average, participants were able to recapitulate the events in the episode in a specific, sequential and highly descriminable way.

\subsection*{The trajectory of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled \citep{Murd62a}. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization \citep{Kaha96, PolyEtal09}. However, these approaches cannot (and were not designed to) capture the rich temporal dynamics inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Video and recall trajectory plots.} \textbf{A.} 2-dimensional embedding of the video events model (reduced with UMAP algorithm). The arrows indicate the forward direction of the video events. \textbf{B.} 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The directional lines connecting the points represent the true video event order. The arrows represent the group-average direction of all recall event transitions (i.e., a line segment connecting two consecutive recall event vectors) that intersected a circle (radius = 0.5) centered on the origin of the arrow. \textbf{C.} Individual trajectory plots for each participant overlayed with the video model (black). The direction of the recall sequence is indicated by the direction of arrows connecting each recall event.  The colors refer to the most likely video event.}
\label{fig:trajectory}
\end{figure}

>>>>

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (e.g.\ every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  For example, some participants' recall topic proportions segmented into just a few events (e.g.\ Participants 1, 4, and 15), while others' recalls segmented into many shorter duration events (e.g.\ Participants 12, 13, and 17).  This suggests that different participants may be recalling the video with different levels of detail-- e.g. some might touch on just the major plot points, whereas others might attempt to recall every minor scene.  The second clear pattern present in every individual participant's recall correlation matrix is that, unlike in the video correlation matrix, there are substantial off-diagonal correlations in participant's recalls.  Whereas each event in the original video (was largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory participants appear to be integrating \textit{across} different events, blending elements of previously recalled and not-yet-recalled events into each newly recalled event~\citep[Figs.~\ref{fig:model}D, \corrmats; also see][]{MannEtal11, HowaEtal12}.

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a common topic model and segmenting the content into events using HMMs.  Next we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector.  This yields a sequence of ``presented'' events from the original movie, and a sequence of (potentially differently ordered) ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[e.g., probability of first recall; Fig.~\listlearning A); ][]{WelcBurn24, PostPhil65, AtkiShif68}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[e.g., lag-conditional response probability; Fig.~\listlearning B; ][]{Kaha96}; and which events they were likely to remember overall~\citep[e.g., serial position recall analyses; Fig.~\listlearning C; ][]{Murd62a}.  In list-learning studies, this set of analyses may be used to gain a nearly complete view into the sequences of recalls participants made~\citep[e.g., ][]{Kaha12}.  However, in naturalistic recall the analyses provide a wholely incomplete picture: it leaves out any attempt to quantify participants' abilities to capture the \textit{content} of what occurred in the video (i.e., their only experimental instruction!).

The dynamic content of the video and participants' recalls is quantified in the corresponding topic proportion matrices.  However, it is difficult to gain deep insights into that content by directly examining the topic proportion matrices (e.g., Figs.~\ref{fig:model}A, D) or the corresponding correlation matrices (Figs.~\ref{fig:model}B, E, \corrmats).  To visualize the time-varying high-dimensional content in a more intuitive way~\citep{HeusEtal18} we projected the topic proportion matrices onto a 2-dimensional space using Uniform Manifold Approximation and Projection~\citep[UMAP; ][]{McInHeal18}.  In this lower-dimensional space, each point represents a single video or recall event, and the distances between the points reflects the distances between their associated topic vectors (Fig.~\ref{fig:trajectory}A, B).

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the video and recalls.}  All panels: the topic proportion matrices have been projected onto a common two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original video (see text), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  \textbf{C.} The recall topic trajectories taken by each individual participant (1--17).  The video's trajectory is shown in black for reference.  (Same format and coloring as Panel A.)}
\label{fig:trajectory}
\end{figure}

@@

% characterize gist of movie and recalls
% x1. movie: fit topic models (add note on dynamic topic models).  fig 1: what do the topics reflect (top words from each topic), example with a few scenes.
% x2. notice block diagonal structure, rather than smooth transitions; motivates HMM approach.  Note about selecting K
% x3. question: do recalls exhibit same structure?  apply topic model to recalls.  note that by projecting everything into a common topic space, the thematic content of each moment of movie/recall can be compared, even though the mix of words used to describe any given event or moment may not be consistent across people (or across annotations/recall)
% x4. recalls exhibit blocky structure, but at different scales.  motivates fitting K separately by person
% x5. match up events (correlations)
% x6. now we can use classic memory analyses to examine structure of recall.  motivate and reference PFR, lag-CRP, and SPC.  but this does not tell us about structure in the *content* of what people remembered, or whether people's recalls are capturing the hierarchical temporal structure of the original movie.
% 7. topic trajectories as a new way of characterizing recall.  assess agreement using rayleigh test.  also examine shapes of each subject's trajectory to show match.  key insight: everyone captures the general shape, but at different "sampling resolutions"
% 8. we can also project back onto the original (word) features to understand more about what any particular part of the trajectory reflects (wordle fig)
% 9. the above analyses tell us about how the trajectory of thematic content in the episode is both preserved and transformed in participants' memories.  we next carried out analyses to understand the neural bases of these phenomena.  analysis 1: which structures are sensitive (during viewing) to the original episode's topic trajectory?  answer: default mode.  anlaysis 2: which structure are sensitive (during viewing) to how the participants will later recall the episode?  answer: vmPFC, default mode.  (potential role of vmPFC in reflecting participant-specific memory transformations and processing)
